Effectiveness of visual test in linear regression diagnositics

### 2022-08-10 paper_comparison/paper_comparison/paper_comparison.tex
update graphs
latest update

The effectiveness of visual inference has already been validated by
\citet{majumder_validation_2013} under relatively simple classical
normal linear regression model settings with only one or two regressors.
Their results suggests visual test is capable of testing the
significance of a single regressor with a similar power as a t-test,
though they expressed that in general it is unnecessary to use visual
inference if there exists a conventional test and they didn't expect the
visual test to perform equally well as the conventional test. In their
third experiment, where there does not exist a proper conventional test,
visual test outperforms the conventional test for a large margin. This
is encouraging as it promotes the use of visual inference in border
field of data science where there are no existing statistical testing
procedures. In fact, lineup protocol has been integrated into some model
diagnostic tools such as \citet{loy2013diagnostic}.

@@

For the purpose of examining the effectiveness of visual test in
regression diagnostics, two experiments were conducted. The first
experiment has ideal scenario for conventional testing, where we would
not expect the visual test to do better than the conventional test. The
second experiment is a scenario where the conventional test is an
approximate test, which visual test may have a chance to match the
performance of the conventional test.

As suggested by many previous studies {[}many studies refs here{]}, it
is reasonable for an subject to evaluate a block of 20 lineups. Subjects
for both two experiments were recruited from Prolific web service {[}ref
here{]}. Data screening procedure was applied during the recruitment,
subjects were required to be fluent in English, with \(98\%\) minimum
approval rate and 10 minimum submissions. During the experiment, they
were asked to select one or more plots that are mostly different from
others, provide a reason for their selections, and express how different
they think the selected plots are from others. If there was no
noticeable difference between plots in a lineup, subject could choose to
select zero plots without providing the reason. Information about
preferred pronoun, age group, education, and previous experience in
visual experiment were also collected. No subject was shown the same
lineup twice.

For each of the experiments, a pool of 12 lineups with obvious visual
patterns were generated. In every block of 20 lineups that presented to
subjects, two different lineups sampled from the corresponding pool of
lineups were included as attention checks. Every subject should get at
least one attention check correct such that the quality of the subject
effort could be ensured.

\hypertarget{non-linear-covariates}{%
\subsection{Non-linear covariates}\label{non-linear-covariates}}

The first experiment is designed to study the ability of human subjects
to detect the effect of a continuous variable \(\boldsymbol{Z}\) which
is a probabilist's Hermite polynomial {[}Herimite ref here{]} of another
variable \(\boldsymbol{X}\) in a two variable statistical model
formulated as:

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{Y} = 1 + \boldsymbol{X} + \boldsymbol{Z} + \boldsymbol{\varepsilon},\\
\boldsymbol{X} = g(\boldsymbol{X}_{raw}, 1), \\
\boldsymbol{Z} = g(\boldsymbol{Z}_{raw}, 1), \\
\boldsymbol{Z}_{raw} = He_j(g(\boldsymbol{X}, 2)),
\end{align}

where \(\boldsymbol{Y}\), \(\boldsymbol{X}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{X}_{raw}\),
\(\boldsymbol{Z}_{raw}\) are \(n\times1\) matrices, \(He_{j}(.)\) is the
\(j\)th-order probabilist's Hermite polynomials,
\(\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\), and
\(g(\boldsymbol{X}, k)\) is a scaling function to enforce the support of
the variable to be \(\{-k, k\}\) defined as

\begin{equation} \label{eq:scaling-function}
g(\boldsymbol{X}, k) = (\boldsymbol{X} - min(\boldsymbol{X}))/max(\boldsymbol{X} - min(\boldsymbol{X})) \times 2k - k, \quad \text{for} \quad k > 0. 
\end{equation}

The null regression model used to fit the data generated by the above
model is formulated as:

\begin{equation} \label{eq:scaling-function}
\boldsymbol{Y} = \beta_0 + \beta_1 \boldsymbol{X} + \boldsymbol{u},
\end{equation}

@@

Clearly, omitted-variable bias will present since the null model leaves
out the higher order term. Data were simulated using four different
order of probabilist's Hermite polynomials (\(j = 2, 3, 6, 18\)), three
different sample sizes (n = 50, 100, 300) , four different standard
deviations of the error (\(\sigma\) = 0.5, 1, 2, 4) and four different
distribution of \(\boldsymbol{X}_{raw}\): (1) \(U(-1, 1)\), (2)
\(N(0, 0.3^2)\), (3) \(lognormal(0, 0.6^2)/3\) and (4) \(u\{1, 5\}\). A
summary of the parameters used in this experiment is given in Table
\ref{tab:parameter-table}.

The values of \(j\) was chosen so that different shapes of non-linearity
are included in the residual plot. These include ``U'' shape, ``S''
shape, ``M'' shape and ``Triple-U'' shape.

@@

The range of \(\sigma\), which is a factor controlling the strength of
the signal, was chosen so that different difficulty levels of lineups
were generated and the estimated power curve would be smooth and
continuous.

@@

Four different distribution were used to generate
\(\boldsymbol{X}_{raw}\). The uniform and the normal distribution are
symmetric and commonly assumed in statistical models. The adjusted
log-normal distribution provides skewed density. And the discrete
uniform distribution provides discreteness in residual plot, which could
enrich the pool of visual patterns.

@@

Three replications are made for each of the parameter values shown in
Table \ref{tab:parameter-table} resulting in 192 different lineups. For
each lineup, the actual data plot was drawn as standard residual plot of
the null model with raw residuals on the y-axis and fitted values on the
x-axis. The 19 null datasets were generated by the residual rotation
technique, and plotted in the same way. The lineup consisted of 20
residual plots with one randomly placed actual data plot. Figure
\ref{fig:example-lineup} is an example of one of these lineups. It was
produced by using \(n = 300\), \(j = 6\), \(\sigma = 0.5\) and
\(X_{raw} \sim N(0.0.3^2)\). The actual data plot location is four. All
five subjects correctly identified the actual data plot for this lineup.

@@

In addition, each lineup is evaluated by five different subjects to
provide reasonable estimates of the visual p-value. Thus,
\(192 \times 3 \times 5 / (20 - 2) = 160\) subjects were recruited to
satisfy the experimental design described above.

@@

A subject was allowed to select zero plots for a lineup, which literally
means there was no visble difference between plots, but the simulated
p-value given in Equation (\ref{eq:p-value-multi}) will effectively drop
the subject from the simulation for this case, leading to inaccurate
estimation of \(p\)-value. Therefore, we treated this case as making one
selection but failing to identify the actual data plot so that Equation
(\ref{eq:p-value-multi}) can be applied correctly.

@@

Table \ref{tab:demographic-table} tabulates the number of subjects,
preferred pronouns, education backgrounds, age groups, and previous
experience in visual experiment. Figure
\ref{fig:demographic-summary-plot} visulizes the marginal distribution
of each of the category. The collated data was a balanced sample among
male and female. Most of the ages of subject were between 18 to 39,
while many of them were between 18 to 24. Very few subjects were awarded
degree higher than Bachelor Degree. Around 40\% of subjects had previous
experience in visual experiment.

@@

For each parameter combination, effect \(E\) is derived from the
Kullback-Leibler divergence (the derivation is provided in {[}appendix
ref here{]}) formulated as:

@@

To study different factors contributing to the power of the visual test,
the same logistic regression model is fit on different subsets of the
collated data grouped by levels of factors. This includes
{[}expansion{]}.

@@

What truly impress us is the huge difference between the estimated power
of visual test and the estimated power of F-test. The margin is largest
at around \(log_e(E) = 2\). An example lineup is included in Figure
\ref{fig:power-overview} where none of subjects detect the actual data
plot positioned at panel 14. It demonstrates that at this level of
difficulty, the designed visual feature is rarely visible, making the
actual data plot indistinguishable from residual plots simulated from
the assumed model. From a communication perspective, given the fact that
the visual difference is unperceivable, the argument that non-linearity
present in the fitted model is less convincing to the public even though
it is true. At around \(log_e(E) = 3\), the margin gets smaller as the
chance of identifying the actual data plot becomes larger. At this level
of difficulty, the designed visual features are usually detectable but
it may not stand out from the lineup as other null plots may happen to
include outliers or visual patterns that are are considered to be more
attractive by human, and thus recognized as the most different plot.
Without knowing the designed visual features beforehand, it is actually
hard to identify the actual data plot by pure image comparison. The
corresponding example lineup for \(log_e(E) = 3\) shown in Figure
\ref{fig:power-overview} has the actual data plot positioned at panel
20, where two out of five subjects detect it. It can be observed that a
M-shape is presented in plot 20, but the signal is not strong enough to
attract all five subjects, resulting in a visual p-value sightly above
the desired significance level \(\alpha = 0.05\). At \(log_e(E) = 4\)
and \(log_e(E) = 6\), the designed visual features become much clear and
attractive, leading to a high percentage of rejection of the null
hypothesis. Figure \ref{fig:power-overview} give example lineups of such
cases.

It is also interesting to observe that the power of visual test behaves
similarly to those inappropriate conventional tests until the designed
visual features really becomes visible

As of shown in Figure \ref{fig:power-overview}, the example lineup for
\(log_e(E) = 3\)

it is of low chance detecting the actual data plot

increase steadily when \(2<log_e(E)<4\). the power of visual test almost
reaches 100\% at \(log_e(E) = 6\).

@@

@@

\includegraphics{paper_comparison_files/figure-latex/unnamed-chunk-9-1.pdf}

>>>>

The effectiveness of visual inference has already been validated by
\citet{majumder_validation_2013} under relatively simple classical
normal linear regression model settings with only one or two regressors.
Their results suggest visual test is capable of testing the significance
of a single regressor with a similar power as a t-test, though they
expressed that in general it is unnecessary to use visual inference if
there exists a conventional test and they didn't expect the visual test
to perform equally well as the conventional test. In their third
experiment, where there does not exist a proper conventional test,
visual test outperforms the conventional test for a large margin. This
is encouraging as it promotes the use of visual inference in border
field of data science where there are no existing statistical testing
procedures. In fact, lineup protocol has been integrated into some model
diagnostic tools such as \citet{loy2013diagnostic}.

@@

For the purpose of examining the effectiveness of visual test in
regression diagnostics, two experiments were conducted. The experiment I
has ideal scenario for conventional testing, where the visual test is
not expected to outperform the conventional test. The experiment II is a
scenario where the conventional test is an approximate test, in which
the visual test may have a chance to match the performance of the
conventional test.

Subjects for both experiments were recruited from an crowdsourcing
platform called Prolific {[}ref here{]}. Prescreening procedure was
applied during the recruitment, subjects were required to be fluent in
English, with \(98\%\) minimum approval rate in other studies and 10
minimum submissions.

During the experiment, every subject was presented with a block of 20
lineups. And for every lineup, the subject was asked to select one or
more plots that are most different from others, provide a reason for
their selections, and evaluate how different they think the selected
plots were from others. If there was no noticeable difference between
plots in a lineup, subjects were permitted to select zero plots without
providing the reason. No subject was shown the same lineup twice.
Information about preferred pronoun, age group, education, and previous
experience in visual experiment were also collected.

A pool of 12 different lineups with obvious visual patterns were
generated for experiment I and experiment II respectively. In every
block of 20 lineups that presented to a subject, two out of 12 lineups
were included as attention checks. A subject's submission was only
accepted if the actual data plot was identified for at least one
attention check. Data of rejected submissions were discarded
automatically to maintain the overall data quality.

\hypertarget{non-linearity}{%
\subsection{Non-linearity}\label{non-linearity}}

Experiment I is designed to study the ability of human subjects to
detect the effect of a random vector \(\boldsymbol{z}\) which is a
probabilist's Hermite polynomial {[}Herimite ref here{]} of another
random vector \(\boldsymbol{x}\) in a two variable statistical model
formulated as:

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{y} = 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} = g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} = g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} = He_j(g(\boldsymbol{z}, 2)),
\end{align}

where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{x}_{raw}\),
\(\boldsymbol{z}_{raw}\) are vector of size \(n\), \(He_{j}(.)\) is the
\(j\)th-order probabilist's Hermite polynomials,
\(\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\), and
\(g(\boldsymbol{x}, k)\) is a scaling function to enforce the support of
the random vector to be \(\{-k, k\}\) defined as

\begin{equation} \label{eq:scaling-function}
g(\boldsymbol{x}, k) = (\boldsymbol{x} - min(\boldsymbol{x}))/max(\boldsymbol{x} - min(\boldsymbol{x})) \times 2k - k, \quad \text{for} \quad k > 0. 
\end{equation}

The null regression model used to fit the realizations generated by the
above model is formulated as:

\begin{equation} \label{eq:scaling-function}
\boldsymbol{y} = \beta_0 + \beta_1 \boldsymbol{x} + \boldsymbol{u},
\end{equation}

@@

Model misspecification presents since the null model leaves out the
higher order term.

Experiment data were simulated using four different order of
probabilist's Hermite polynomials (\(j = 2, 3, 6, 18\)), three different
sample sizes (n = 50, 100, 300) , four different standard deviations of
the error (\(\sigma\) = 0.5, 1, 2, 4) and four different distribution of
\(X_{raw}\): (1) \(U(-1, 1)\), (2) \(N(0, 0.3^2)\), (3)
\(lognormal(0, 0.6^2)/3\) and (4) \(u\{1, 5\}\). A summary of the
parameters used in this experiment is given in Table
\ref{tab:parameter-table}.

The values of \(j\) was chosen so that different shapes of non-linearity
were included in the residual plot. These include ``U'' shape, ``S''
shape, ``M'' shape and ``Triple-U'' shape.

@@

The range of \(\sigma\), which is a factor controlling the strength of
the signal, was chosen so that different difficulty levels of lineups
were generated, and therefore, the estimated power curve would be smooth
and continuous.

@@

Four different distribution were used to generate \(X_{raw}\). The
uniform and the normal distribution are symmetric and commonly assumed
in statistical models. The adjusted log-normal distribution provides
skewed density. And the discrete uniform distribution provides
discreteness in residual plot, which could enrich the pool of visual
patterns.

@@

Three replications are made for each of the parameter values shown in
Table \ref{tab:parameter-table} resulting in 192 different lineups. For
each lineup, the actual data plot was drawn as a standard residual plot
of the null model with raw residuals on the y-axis and fitted values on
the x-axis. The 19 null datasets were generated by the residual rotation
technique, and plotted in the same way. The lineup consisted of 20
residual plots with one randomly placed actual data plot. Figure
\ref{fig:example-lineup} is an example of one of these lineups. It was
produced by using \(n = 300\), \(j = 6\), \(\sigma = 0.5\) and
\(X_{raw} \sim N(0.0.3^2)\). The actual data plot location was four. All
five subjects correctly identified the actual data plot for this lineup.

@@

In addition, each lineup is designed to be evaluated by five different
subjects to provide reasonable estimates of the visual p-value. Thus,
\(192 \times 3 \times 5 / (20 - 2) = 160\) subjects were recruited to
satisfy the design of the experiment I.

@@

A subject was allowed to select zero plots for a lineup if there was no
visible difference between plots, but the simulated visual p-value given
in Equation (\ref{eq:p-value-multi}) will effectively drop the subject
from the simulation for this case, leading to inaccurate estimation of
\(p\)-value. Therefore, we treated this case as making one selection but
failing to identify the actual data plot so that Equation
(\ref{eq:p-value-multi}) can be applied correctly.

@@

Table \ref{tab:demographic-table} tabulates the number of subjects,
preferred pronouns, education backgrounds, age groups, and previous
experience in visual experiment. Figure
\ref{fig:demographic-summary-plot} visualizes the marginal distribution
of each of the category. The collated data was a balanced sample among
male and female. Most of the ages of subject were between 18 to 39,
while many of them were between 18 to 24. Very few subjects were awarded
degree higher than Bachelor Degree. Around 40\% of subjects had previous
experience in visual experiment.

@@

For each parameter combination, effect \(E\) is derived from the
Kullback-Leibler divergence (see {[}appendix ref here{]}) formulated as:

@@

To study various factors contributing to the power of the visual test,
the same logistic regression model is fit on different subsets of the
collated data grouped by levels of factors. This includes
{[}expansion{]}.

@@

What truly impress us is the huge difference between the estimated power
of visual test and the estimated power of F-test. The margin is largest
at around \(log_e(E) = 2\). An example lineup is included in Figure
\ref{fig:power-overview} where none of subjects detect the actual data
plot positioned at panel 14. It demonstrates that at this level of
difficulty, the designed visual feature is rarely visible, making the
actual data plot indistinguishable from residual plots simulated from
the assumed model. From a communication perspective, given the fact that
the visual difference is unperceivable, the argument that non-linearity
present in the fitted model is less convincing to the public even though
it is true. At around \(log_e(E) = 3\), the margin gets smaller as the
chance of identifying the actual data plot becomes larger. At this level
of difficulty, the designed visual features are usually detectable but
it may not stand out from the lineup as other null plots may happen to
include outliers or visual patterns that are are considered to be more
attractive by human, and thus recognized as the most different plot.
Without knowing the designed visual features beforehand, it is actually
hard to identify the actual data plot by pure image comparison. The
corresponding example lineup for \(log_e(E) = 3\) shown in Figure
\ref{fig:power-overview} has the actual data plot positioned at panel
20, where two out of five subjects detect it. It can be observed that a
M-shape is presented in plot 20, but the signal is not strong enough to
attract all five subjects, resulting in a visual p-value sightly above
the desired significance level \(\alpha = 0.05\). At \(log_e(E) = 4\)
and \(log_e(E) = 6\), the designed visual features become much clear and
attractive, leading to a high percentage of rejection of the null
hypothesis. Figure \ref{fig:power-overview} gives example lineups of
such cases.

@@

\hypertarget{distributions-of-regressor}{%
\subsubsection{Distributions of
regressor}\label{distributions-of-regressor}}

The impact of the distribution of \(X_raw\) on the power is shown in
Figure \ref{fig:dist-power}. The power curve of F-test is stable across
different distributions, while the visual test has a steeper power curve
for normal and uniform distribution. BP-test performs worse for discrete
uniform distribution and uniform distribution but has relatively high
power for normal distribution. SW-test outperforms BP-test for discrete
uniform distribution but remains as the worst test for other
distributions. The results indicate those inappropriate residual-based
tests are sensitive to the distribution of the regressor.

\begin{figure}
\centering
\includegraphics{paper_comparison_files/figure-latex/dist-power-1.pdf}
\caption{\label{dist-power}}
\end{figure}

\hypertarget{shpaes-of-non-linearity}{%
\subsubsection{Shpaes of non-linearity}\label{shpaes-of-non-linearity}}

Figure \ref{fig:shape-power} illustrates the change of power under
different shape of non-linearity. Similar to the power curves shown in
\ref{fig:dist-power}, F-test is stable under different shapes. The power
curve of visual test also behaves similarly across different shapes.
What vary are the power curve of BP-test and SW-test. For Triple-U
shape, both BP-test and SW-test are insensitive to the change of the
effect. And for W shape, both tests have almost identical power curves.
It can be observed that both tests performs the best for U-shape.

\begin{figure}
\centering
\includegraphics{paper_comparison_files/figure-latex/shape-power-1.pdf}
\caption{\label{shape-power}}
\end{figure}

\hypertarget{reasons-for-making-selections}{%
\subsubsection{Reasons for making
selections}\label{reasons-for-making-selections}}

\includegraphics{paper_comparison_files/figure-latex/unnamed-chunk-12-1.pdf}

\hypertarget{heteroskedasticity}{%
\subsection{Heteroskedasticity}\label{heteroskedasticity}}

\includegraphics{paper_comparison_files/figure-latex/unnamed-chunk-15-1.pdf}

\includegraphics{paper_comparison_files/figure-latex/unnamed-chunk-16-1.pdf}

\includegraphics{paper_comparison_files/figure-latex/unnamed-chunk-17-1.pdf}

\includegraphics{paper_comparison_files/figure-latex/unnamed-chunk-18-1.pdf}

@@

\includegraphics{paper_comparison_files/figure-latex/unnamed-chunk-19-1.pdf}
